Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-02 _tex/index.tex
Built site for gh-pages

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these designs: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(or mean), referred to as the \emph{modal discriminal process},
indicates the stimulus' position on the trait continuum, and its
dispersion, known as the \emph{discriminal dispersion}, reflects the
variability in the perceived trait level of the stimulus.

@@

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution. The
mode (or mean) of this distribution, representing the (average) relative
separation, is given by the difference between the modal discriminal
processes of the stimuli \(S_{BA} = S_{B} - S_{A}\). Meanwhile, the
dispersion of the distribution, reflecting the variability in the
relative separation, is calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
their discriminal processes. This correlation quantifies how much the
judge's perception of the trait in one stimulus influences his
perception of the same trait in the other.

@@

Additionally, Figure~\ref{fig-correlation} illustrates how varying
correlations influence the distribution of the discriminal difference
for the same two texts. The figure shows that as the correlation
increases, indicating a stronger influence of the judge's perception of
quality in one text on his perception of the other, the distribution of
the discriminal difference becomes narrower. This narrowing affects the
area under the curve that determines the comparison outcome and,
ultimately, the conclusions drawn from this outcome.

For instance, when two texts clearly differ in quality, as depicted in
Figure~\ref{fig-discriminal_process}, higher correlations increase the
likelihood that the discriminal difference more distinctly favors text B
over text A, as indicated by the shaded gray area under the curve in
Figure~\ref{fig-correlation}. Conversely, when the two texts have
similar or identical quality levels, higher correlations reduce the
likelihood that the discriminal difference distinctly favors one text
over the other, as the distribution narrows around zero (not depicted in
the figure).

>>>>

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(or the mean), referred to as the \emph{modal discriminal process},
indicates the stimulus' position on the trait continuum, and its
dispersion, known as the \emph{discriminal dispersion}, reflects the
variability in the perceived trait level of the stimulus.

@@

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution. The
mode (or the mean) of this distribution, representing the (average)
relative separation, is given by the difference between the modal
discriminal processes of the stimuli \(S_{BA} = S_{B} - S_{A}\).
Meanwhile, the dispersion of the distribution, reflecting the
variability in the relative separation, is calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
the discriminal processes of the stimuli. This correlation quantifies
the dependence of the judge's perception of the trait in one stimulus on
his perception of the same trait in the other.

@@

Notably, the correlation between the discriminal processes, \(\rho\),
plays a pivotal role in determining comparison outcomes by shaping the
distribution of the discriminal difference between the stimuli.
Specifically, as the correlation increases, reflecting a stronger
dependence of the judge's perception of quality in one stimulus on his
perception of the other, the distribution of the discriminal difference
narrows. This narrowing ultimately impacts the area under the curve that
determines the comparison outcome and, consequently, the conclusions
drawn from the comparison.

Figure~\ref{fig-correlation} illustrates how varying correlations
influence the distribution of the discriminal difference for the two
texts depicted in Figure~\ref{fig-discriminal_process}. Since the texts
differ in quality, higher correlations increase the likelihood that the
discriminal difference distinctly favors text B over text A. This is
evident from the larger proportion of the area under the curve,
\(P(B > A)\), that lies above zero (shaded gray area in
Figure~\ref{fig-correlation}). Conversely, it is not hard to infer that
if the texts had similar or identical quality levels, higher
correlations would likely reduce the chance that the discriminal
difference distinctly favors one text over the other. This probability
reduction occurs because the distribution of the discriminal difference
would become more narrowly centered around zero (not illustrated).
